Title Goes Here

### 2020-02-24 paper/paper.tex
Expanding on new representation in paper

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall]{acmart}
\usepackage{booktabs}
\usepackage{listings}

@@

@@

@@

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmMonth{1}

@@

% \authornotemark[1]
% \email{webmaster@marysville-ohio.com}
% \affiliation{%
%   \institution{Institute for Clarity in Documentation}
%   \streetaddress{P.O. Box 1212}
%   \city{Dublin}
%   \state{Ohio}
%   \postcode{43017-6221}
% }

% \author{Lars Th{\o}rv{\"a}ld}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{larst@affiliation.org}

% \author{Valerie B\'eranger}
% \affiliation{%
%   \institution{Inria Paris-Rocquencourt}
%   \city{Rocquencourt}
%   \country{France}
% }

% \author{Aparna Patel}
% \affiliation{%
%  \institution{Rajiv Gandhi University}
%  \streetaddress{Rono-Hills}
%  \city{Doimukh}
%  \state{Arunachal Pradesh}
%  \country{India}}

% \author{Huifen Chan}
% \affiliation{%
%   \institution{Tsinghua University}
%   \streetaddress{30 Shuangqing Rd}
%   \city{Haidian Qu}
%   \state{Beijing Shi}
%   \country{China}}

% \author{Charles Palmer}
% \affiliation{%
%   \institution{Palmer Research Laboratories}
%   \streetaddress{8600 Datapoint Drive}
%   \city{San Antonio}
%   \state{Texas}
%   \postcode{78229}}
% \email{cpalmer@prl.com}

% \author{John Smith}
% \affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
% \email{jsmith@affiliation.org}

% \author{Julius P. Kumquat}
% \affiliation{\institution{The Kumquat Consortium}}
% \email{jpkumquat@consortium.net}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Domain specific languages (DSLs) provide a powerful tool for to abstract over
  a class of problems. In this paper, we introduce a technique for representing
  algebraic data types and pattern matching in a DSL embedded in Haskell. This
  representation is automated through the use of GHC Generics and Template Haskell.
\end{abstract}

@@

@@

example' :: Int
example' =
  abs (CaseExp (rep x)
               (SumMatch (OneProdMatch (\c -> ord c))
                         (OneSumMatch
                           (OneProdMatch (\i -> i)))))
\end{lstlisting}

@@

@@

\begin{lstlisting}
instance (ERep a, ERep b) => ERep (Either a b) where
  type ERepTy (Either a b) = Either a b

@@

instance (ERep a, ERep b) => ERep (a, b) where
  type ERepTy (a, b) = (a, b)

@@

\begin{lstlisting}
data E t where
  ...
  CaseExp :: ERep t => E t -> SumMatch (ERepTy t) r -> E r
  ...
\end{lstlisting}

A value of type \ttt{SumMatch a b} represents a computation within the EDSL
which destructures a value of type \ttt{E a} and produces a value of type \ttt{E b}.
Therefore, \ttt{SumMatch (ERepTy t) r} represents a computation which destructures
a value of type \ttt{E (ERepTy t)} and produces a value of type \ttt{E b}.

\subsection{\ttt{SumMatch}}

The overall structure of \ttt{SumMatch} is a (heterogeneous) list of \ttt{ProdMatch} values.
Each item of this list corresponds exactly to one branch in the original \ttt{case} match.

\begin{lstlisting}
data SumMatch s t where
  SumMach ::
    (ERep a, ERep b, ERepTy b ~ b)
      => ProdMatch a r -> SumMatch b r -> SumMatch (Either a b) r

  EmptyMatch :: SumMatch Void r

  OneSumMatch ::
    (ERep a, ERep b, ERepTy a ~ a)
      => ProdMatch a b -> SumMatch a b
\end{lstlisting}

Note the \ttt{ERepTy a ~ a} constraints. This constraint ensures that the type \ttt{a}
is already in canonical form (that is, consists entirely of \ttt{Either}, \ttt{(,)} and
base types).

\subsection{\ttt{ProdMatch}}

\ttt{ProdMatch s t} is equivalent to a function a curried function from \ttt{s} to \ttt{t}.
Note that \ttt{s} will be a (potentially nested) pair type. For example,
\ttt{ProdMatch (a, (b, c)) r} will be equivalent to \ttt{a -> b -> c -> r}.

\begin{lstlisting}
data ProdMatch s t where
  ProdMatch ::
    (ERep a, ERep b)
      => (E a -> ProdMatch b r) -> ProdMatch (a, b) r

  OneProdMatch :: ERep a => (E a -> E r) -> ProdMatch a r

  NullaryMatch :: ERep r => E r -> ProdMatch a r
\end{lstlisting}

\section{Constructing values}

\begin{lstlisting}
construct ::
  Construct (a -> b)
    => (a -> b) -> LiftedFn (a -> b)
construct = construct' . Construct

-- Example: LiftedFn (a -> b -> c) is: E a -> E b -> E c
type family LiftedFn t where
  LiftedFn (a -> b) = E a -> LiftedFn b
  LiftedFn b = E b

class Construct t where
  construct' :: E t -> LiftedFn t
\end{lstlisting}

Example GHCi session, demonstrating \ttt{construct}:

\begin{lstlisting}
ghci> :t construct (,)
construct (,)
  :: (ERep a, ERep b) => E a -> E b -> E (a, b)

ghci> :t construct (,,)
construct (,,)
  :: (ERep a, ERep b, ERep c) =>
     E a -> E b -> E c -> E (a, b, c)

ghci> :t construct (,,) (rep False) (rep 3.1 :: E Float) (rep 1 :: E Int)
construct (,,) (rep False) (rep 3.1 :: E Float) (rep 1 :: E Int)
  :: E (Bool, Float, Int)
\end{lstlisting}

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen]{acmart}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{tcolorbox}

@@

\newcommand{\typeeq}{\raise.17ex\hbox{$\scriptstyle\mathtt{\sim}$}\,}

@@

% \newenvironment<>{todo}[1][]{
%     \setbeamercolor{block body example}{fg=black,bg=blue}
%     \setbeamercolor{block title example}{fg=white,bg=red!75!black}
%     \setbeamertemplate{blocks}[rounded][shadow=false]
%   \begin{todo}[]}{\end{todo}
% }

@@

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Domain specific languages (DSLs) provide a powerful tool for to abstract over
  a class of problems. In this paper, we introduce a technique for representing
  algebraic data types and pattern matching in a DSL embedded in Haskell. This
  representation is mechanized through the use of GHC Generics and and GHC Core plugin.
\end{abstract}

@@

\subsection{Overview}

In this EDSL, the type representing values in the expression language is \ttt{E
t} and there are two basic functions which translate values to and from this
expression language with the following type signatures:

\begin{lstlisting}
rep :: a -> E a
abs :: E a -> a
\end{lstlisting}

In this paper, the data constructors of the \ttt{E} type will be described
incrementally as they are needed.

@@

example' :: Int
example' =
  abs (CaseExp
        (rep x)
        (SumMatchExp (OneProdMatch
                       (Lam 1 (Ord (Var 1))))
                     (OneSumMatch
                       (OneProdMatch
                         (Lam 2 (Var 2)))))
\end{lstlisting}

@@

Briefly, a algebraic type \ttt{T} with an automatically generated \ttt{ERep}
instance will be given representation in terms of \ttt{Either}, \ttt{(,)},
\ttt{()} and \ttt{Void}. This "standardized" type representation is
given by \ttt{ERepTy T}. \ttt{ERepTy} is an type family associated to
the type class \ttt{ERep}. This type will be called the \textit{canonical type}
of \ttt{T}.

@@

\begin{lstlisting}
instance (ERep a, ERep b)
  => ERep (Either a b) where

  type ERepTy (Either a b) = Either a b

@@

instance (ERep a, ERep b)
  => ERep (a, b) where

  type ERepTy (a, b) = (a, b)

@@

\begin{lstlisting}
data E t where
  ...
  CaseExp ::
    (ERep t, ERepTy (ERepTy t) ~ ERepTy t)
      => ERep t
         -> E (SumMatch (ERepTy t) r)
         -> E r
  ...
\end{lstlisting}

The equality constraint ensures that a canonical type is its own canonical type.

\begin{tcolorbox}
  \textbf{TODO}: Should we make a synonym for this constraint since it is used in
multiple places? Like this:
\begin{lstlisting}
type CanonicalIsIdem t =
  ERepTy (ERepTy t) ~ ERepTy t
\end{lstlisting}
\end{tcolorbox}

(The above code block is a slight simplification of the actual implementation,
which has an additional type variable which is used at an "intermediate" point.
The expanded form is in place to simplify the Core transformation.)

The type \ttt{SumMatch} is defined as

\begin{lstlisting}
newtype SumMatch a b =
  MkSumMatch { runSumMatch :: a -> b }
\end{lstlisting}

For the moment, we will primarily use this type as a type tag and ignore
the values it can take on.

A value of type \ttt{E (SumMatch a b)} represents a computation within the EDSL
which destructures a value of type \ttt{E a} and produces a value of type \ttt{E b}.
Therefore, \ttt{E (SumMatch (ERepTy t) r)} represents a computation which destructures
a value of type \ttt{E (ERepTy t)} and produces a value of type \ttt{E b}.

\subsection{\ttt{E (SumMatch a b)}}

The overall structure of a \ttt{E (SumMatch a b)} value is a (heterogeneous)
list of \ttt{E (ProdMatch x y)} values.  Each item of this list corresponds
exactly to one branch in the original \ttt{case} match.

The following constructors generate \ttt{SumMatch}-tagged values in the
expression language:

\begin{lstlisting}
data E t where
  ...
  SumMatchExp ::
    (ERep a, ERep b, ERepTy b ~ b)
      => E (ProdMatch a r)
         -> E (SumMatch b r)
         -> E (SumMatch (Either a b) r)

  OneSumMatch ::
    (ERep a, ERep b, ERepTy a ~ a)
      => E (ProdMatch a b)
         -> E (SumMatch a b)

  EmptyMatch ::
    (ERep b) => E (SumMatch Void b)
  ...
\end{lstlisting}

Note the \ttt{ERepTy a \typeeq a} constraints. This constraint ensures that the type
\ttt{a} is already in canonical form (that is, consists entirely of
\ttt{Either}, \ttt{(,)} and base types).

\subsection{\ttt{E (ProdMatch s t)}}

\ttt{E (ProdMatch x y)} is equivalent to a curried function from \ttt{x} to
\ttt{y} in the expression language .  Note that \ttt{s} will be a (potentially
nested) pair type. For example, \ttt{E (ProdMatch (a, (b, c)) r)} will be
equivalent to \ttt{E (a -> b -> c -> r)}.

\begin{lstlisting}
data E t where
  ...
  ProdMatchExp ::
    (ERep a, ERep b)
      => E (a -> ProdMatch b r)
         -> E (ProdMatch (a, b) r)

  NullaryMatch ::
    (ERep a)
      => E r -> E (ProdMatch a r)

  OneProdMatch ::
    (ERep a)
      => E (a -> b) -> E (ProdMatch a b)
  ...
\end{lstlisting}

The \ttt{ProdMatch} type is defined similarly to the \ttt{SumMatch} type and is
similiarly used primarily as a type tag:

\begin{lstlisting}
newtype ProdMatch a b =
  MkProdMatch { runProdMatch :: a -> b }
\end{lstlisting}

\subsection{Connection to Church encodings}
TODO: Write this subsection

\subsection{An aside on \ttt{ProdMatch} and \ttt{SumMatch} values}

Though \ttt{ProdMatch} and \ttt{SumMatch} are used throughout this EDSL as type
tags, they do have values and they are not trivial values. The reason for this
is that it connects the encoded pattern matches (values from the \ttt{E}) type
to their semantics. A value of type \ttt{E (SumMatch a b)} is an expression
which takes in a value of type \ttt{a} (embedded within the expression
language), internally performs some pattern matching, and produces a value of
type \ttt{b} (again, embedded within the expression language). This is exactly
the semantics of a function from \ttt{a} to \ttt{b}. Likewise for \ttt{ProdMatch a b}.

Recall the previously mentioned function

\begin{lstlisting}
abs :: E a -> a
\end{lstlisting}

Now consider at the type of \ttt{abs} when it is specialized to take \ttt{E
(SumMatch a b)} values:

\begin{lstlisting}
abs :: E (SumMatch a b) -> SumMatch a b
\end{lstlisting}

The \ttt{SumMatch a b} value which \ttt{abs} returns is exactly the function
which pattern matches according to its input value.

Likewise for \ttt{ProdMatch a b} values.

\section{Representing lambdas}
This representation of pattern matching depends on the ability to bring
function values into the expression language. This is accomplished
with the following constructors:

\begin{lstlisting}
data E t where
  ...
  Lam ::
    (ERep a, Typeable a)
      => Name a -> E b -> E (a -> b)

  Var :: (Typeable a) => Name a -> E a
\end{lstlisting}

The \ttt{Typeable} constraints are necessary to lookup correctly typed
values in the variable binding environment later on.

The \ttt{Name t} type represents a lambda variable identifier together with its
type \ttt{t}:

\begin{lstlisting}
newtype Name a = Name Int
  deriving (Eq, Show)

namesEq :: forall a b.
  (Typeable a, Typeable b)
    => Name a -> Name b -> Maybe (a :~: b)
namesEq (Name n) (Name n') =
  case eqT :: Maybe (a :~: b) of
    Just Refl
      | n == n' -> Just Refl
      | otherwise -> Nothing
    Nothing   -> Nothing
\end{lstlisting}

(Note that the \ttt{ScopedTypeVariables} extension is enabled.)

In the Core transformation, each lambda is given a \ttt{Name} with a globally
unique \ttt{Int}, sidestepping any name capture issues.

The following datatypes are used to represent a variable binding environment
of typed names to expression language values:

\begin{lstlisting}
data EnvMapping where
  (:=>) :: forall a.
    Typeable a
    => Name a -> E a -> EnvMapping
\end{lstlisting}

This type encodes a single variable binding, with values of the form \ttt{n :=>
v}, where \ttt{n} is a typed name and \ttt{v} is the value it is bound to.

These bindings are grouped together in the \ttt{Env} type:

\begin{lstlisting}
newtype Env = Env [EnvMapping]

emptyEnv :: Env
emptyEnv = Env []

extendEnv :: Env -> EnvMapping -> Env
extendEnv (Env maps) m = Env (m:maps)

envLookup :: Typeable a
  => Env -> Name a -> Maybe (E a)
envLookup (Env maps) = go maps
  where
    go []                = Nothing

    go ((n' :=> e):rest) =
      case namesEq n n' of
        Just Refl -> Just e
        Nothing   -> go rest
\end{lstlisting}

\section{Recovering standard semantics for pattern matches}

\begin{lstlisting}
abs :: forall t. E t -> t
abs = absEnv emptyEnv

sumMatchAbs ::
  (ERepTy (ERepTy s) ~ ERepTy s, ERep s)
  => Env
    -> E (SumMatch (ERepTy s) t)
    -> s
    -> t
...

prodMatchAbs :: (ERep s)
  => Env
     -> E (ProdMatch s t)
     -> s
     -> t
...

absEnv :: forall t. Env -> E t -> t
absEnv env (CaseExp x f) =
  sumMatchAbs f (absEnv env x)

absEnv env (LeftExp x) =
  Left (absEnv env x)

absEnv env (RightExp y) =
  Right (absEnv env y)

absEnv env (PairExp x y) =
  (absEnv env x, absEnv env y)

absEnv env (Repped x) = unrep x

absEnv env env (Var v) =
  case envLookup env v of
    Just x -> absEnv env
    Nothing ->
      error
        ("No binding for name "
          ++ show v)

...
\end{lstlisting}
